---
title: "Assignment 09: Data Scraping"
author: "Enikoe Bihari"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

# check working directory
getwd()

# import libraries
library(tidyverse)
# install.packages("rvest")
library(rvest)
library(lubridate)
# install.packages("cowplot")
library(cowplot)
# install.packages("gridExtra")           
library("gridExtra")  
# install.packages("grid")                          
library("grid")
library(ggplot2)

# create a theme with gray defaults
theme1 <- theme_gray(base_size = 12) +
  theme(axis.text = element_text(color = "grey50"), 
        legend.position = "top",
        axis.title = element_text(color = "grey20"),
        legend.key.width = unit(2, "cm"))

# set it as the default theme
theme_set(theme1)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2021 to 2020 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2

# create a webpage object from the url
Durham_LWSP_page <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
Durham_LWSP_page

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Max Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3

# get water system name from the webage as text
water.system.name <- Durham_LWSP_page %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

# get PWSID
pwsid <- Durham_LWSP_page %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid

# get ownership
ownership <- Durham_LWSP_page %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership
  
# get max withdrawals
max.withdrawals.mgd <- Durham_LWSP_page %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```

4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4

# create a dataframe
df_mgd <- data.frame("month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                      "year" = rep(2020,12),
                      "max.withdrawals.mgd" = as.numeric(max.withdrawals.mgd))

# put in other attributes and the date
df_mgd <- df_mgd %>% 
  mutate(water.system.name = !!water.system.name,
         pwsid = !!pwsid,
         ownership = !!ownership,
         date = my(paste(month,"-",year)))
           
#5

# plot the data
ggplot(df_mgd, aes(x=date, y=max.withdrawals.mgd)) + 
  geom_line() + 
  geom_smooth(method="loess", se=FALSE, color = "mediumorchid") +
  labs(title = "Maximum Daily Water Use in Durham in 2020",
       y="Withdrawal (mgd)",
       x="Date") 

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

# define all the inputs to the function
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
the_PWSID <- '03-32-010'
the_year <- 2020
the_scrape_url <- paste0(the_base_url, "pwsid=", the_PWSID, "&year=", the_year)
print(the_scrape_url)

# define the function
scrape.mgd <- function(the_year, the_PWSID, the_base_url){
  
  # access the website
  the_website <- read_html(paste0(the_base_url, "pwsid=", the_PWSID, "&year=", the_year))
  # print(the_website)
  
  # Set the element address variables
  the_watersytemname_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  the_pwsid_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  the_ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_mgd_tag <- 'th~ td+ td'
  
  # scrape the data items
  the_watersystem <- the_website %>% html_nodes(the_watersytemname_tag) %>% html_text()
  the_pwsid <- the_website %>%   html_nodes(the_pwsid_tag) %>%  html_text()
  the_ownership <- the_website %>% html_nodes(the_ownership_tag) %>% html_text()
  max_mgd <- the_website %>% html_nodes(the_mgd_tag) %>% html_text()
  
  # create a dataframe
  the_df <- data.frame("month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                        "year" = rep(the_year,12),
                        "max.withdrawals.mgd" = as.numeric(max_mgd)) %>% 
    mutate(water.system.name = !!the_watersystem,
           pwsid = !!the_pwsid,
           ownership = !!the_ownership,
           date = my(paste(month,"-",year)))
  
  # pause
  #Sys.sleep(1) #uncomment this if you are doing bulk scraping!
  
  # return the dataframe
  return(the_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

# define all the new inputs to the function
the_PWSID_durham <- '03-32-010'
the_year_2015 <- 2015

# run the function
df_mgd_durham_2015 = scrape.mgd(the_year_2015, the_PWSID_durham, the_base_url)

# plot the data
ggplot(df_mgd_durham_2015, aes(x=date, y=max.withdrawals.mgd)) + 
  geom_line() + 
  geom_smooth(method="loess", se=FALSE, color = "cyan3") +
  labs(title = "Maximum Daily Water Use in Durham in 2015",
       y="Withdrawal (mgd)",
       x="Date") 

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.Asheville.2015.data}
#8

# define all the new inputs to the function
the_PWSID_asheville <- '01-11-010'
the_year_2015 <- 2015

# run the function
df_mgd_asheville_2015 = scrape.mgd(the_year_2015, the_PWSID_asheville, the_base_url)

# plot the data
ggplot() +
  geom_line(data = df_mgd_durham_2015, 
            aes(x=date, y=max.withdrawals.mgd)) +
  geom_smooth(data = df_mgd_durham_2015, 
              aes(x=date, y=max.withdrawals.mgd), 
              method="loess", 
              se=FALSE, 
              color = "seagreen3") +
  geom_line(data = df_mgd_asheville_2015, aes(x=date, y=max.withdrawals.mgd)) +
  geom_smooth(data = df_mgd_asheville_2015, 
              aes(x=date, y=max.withdrawals.mgd), 
              method="loess", 
              se=FALSE, 
              color = "indianred1") +
  labs(title = "Maximum Daily Water Use in Durham vs. Asheville in 2015",
       y="Withdrawal (mgd)",
       x="Date") +
  annotate("text", 
           x = as.Date("2/1/2015", format = "%m/%d/%Y"), 
           y = 52, 
           label = "Durham",
           color = "seagreen3",
           cex = 5) +
  annotate("text", 
           x = as.Date("2/1/2015", format = "%m/%d/%Y"), 
           y = 27, 
           label = "Asheville",
           color = "indianred1",
           cex = 5)

```

9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2020.Add a smoothed line to the plot.

```{r}
#9

# define all the new inputs to the function
the_PWSID_asheville <- '01-11-010'
the_years_2010.2019 <- rep(2010:2020)

# run the function with lapply
the_dfs_asheville <- lapply(X = the_years_2010.2019,
                  FUN = scrape.mgd,
                  the_PWSID=the_PWSID_asheville,
                  the_base_url = the_base_url)

# put into a single dataframe
the_df_asheville <- bind_rows(the_dfs_asheville)

# plot the data
ggplot(the_df_asheville, aes(x=date, y=max.withdrawals.mgd)) + 
  geom_line() + 
  geom_smooth(method="loess", se=FALSE, color = "slateblue2") +
  labs(title = "Maximum Daily Water Use in Asheville, 2010-2020",
       y="Withdrawal (mgd)",
       x="Date") 

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>Water usage in Asheville seemed to dip around 2015, but has been increasing since then (with 2020 water usage levels being higher than they were in 2010 before the dip).